---
title: "Shiny education"
author: "Owen"
date: "22/03/2022"
output: 
  github_document
always_allow_html: yes
---

\newpage

# Why shiny?

Do you ever want to give a non-programmer the ability to investigate a data-set themselves? Sometimes sharing insights with decision makers can be made more persuasive if the decision makers can do some of the investigations themselves.

*R Shiny* bridges the gap between an analyst who is working in R and a decision maker who doesn't program but want to explore the data. A Shiny application uses standard R analytic programming and then builds an HTLM interface that can be interacted with through a web browser.

# Quick start

Let's build a Shiny application that addresses the problem **"What are the biggest schools by enrollment for each school type?"**.

## A non-shiny approach
If we were not using Shiny we could provide a static image such as this:

```{r fig.width=7, fig.height=7, message=FALSE, warning=FALSE}
library(readr)
library(tidyverse)

X8_sa_enrolment_by_school <- read_csv("https://data.sa.gov.au/data/dataset/d385385a-6d79-4ef7-ab0f-0cfc583469f6/resource/7223f681-d340-4fc9-b92b-403d158f521f/download/8-sa-enrolment-by-school-.csv")

X8_sa_enrolment_by_school %>%
  group_by(School_Type) %>%
  slice_max(order_by = Year_2021, n = 6) %>%
  ungroup() %>%
  ggplot(aes(x = Year_2021, 
             y = fct_reorder(School_Name, Year_2021), 
             fill = School_Type)) + 
  geom_col(show.legend = FALSE) + 
  facet_wrap(~School_Type, 
             scales = "free", 
             ncol = 1) + 
  labs(title = "Biggest schools for each school type", x = "Enrollemnt in 2021", y = NULL)
```

But what if we could give decision makers an interactive webpage where they could select the school type and the plot would render to show what they had selected?

\newpage

## Anatomy of a shiny app

If you open RStudio and go to *File > New File > Shiny Web App* you can name your new application. This actually creates a new directory with the name you chose. In that directory is a standard R file called `app.R`. It has an example Shiny app already written there to show you how to get started.

### A tale of two code sections

There are two parts to every `app.R` file: `ui` and `server`. The `ui` sets out how the user interface will look and feel and defines the objects that the user will select. The `server` contains all the R analysis that is done in the background. It is written in the form of a R function (within curly brackets) and contains at least one expression that is also within curly brackets which *reacts* to the user's input.

### ui

The layout we might want could include three sections:

* Title
* User interface
* Resultant plot

```{r eval=FALSE}
ui <- fluidPage(
  
  titlePanel(),
  
  sidebarLayout(
    sidebarPanel(
    )
  ),
  
  mainPanel(
  )
)
```

We can then fill in each section:

* Title: A string in the `titlePanel()` section
* User interface: Some `radioButtons()` to select one of the levels of `X8_sa_enrolment_by_school$School_Type` in the `sidebarPanel()` section.
* Resultant plot: A named `plotOutput()` in the `mainPanel()` section.

\newpage

```{r eval=FALSE}
ui <- fluidPage(

    titlePanel("Biggest schools by enrollment"),

    # Sidebar with a slider input for number of bins 
    sidebarLayout(
        sidebarPanel(
          radioButtons("selected_type",
                        "Type of school",
                       unique(X8_sa_enrolment_by_school$School_Type)
          )
        ),

        mainPanel(
           plotOutput("biggestPlot")
        )
    )
)
```

### server

The `server` section is really just a R function

```{r eval=FALSE}
server <- function(input, output) {

}
```

The only thing we have to put in here is the code that creates `biggestPlot` which we are calling back in the `ui` section. And as part of that code, we have to let the plot be updated based on the user's selection of the `selected_type`.

If we were just creating a plot to show one `School_Type` we could use `filter()` like this:

```{r}
      X8_sa_enrolment_by_school %>%
        filter(School_Type == "Primary Education") %>%
        slice_max(order_by = Year_2021, n = 12) %>%
        ggplot(aes(x = Year_2021, 
                   y = reorder(School_Name, Year_2021), 
                   fill = School_Type)) + 
        geom_col(show.legend = FALSE) + 
  labs(title = "Biggest schools 2021", x = "Enrollments 2021", y = NULL)
```

#### Updating from inputs
So all we have to do is replace the `"Primary Education"` with the user's selection of `selected_type` in the `ui`.

When it is run, Shiny creates a temporary object called `input` which stores the currently selected values of all the user inputs. We access them using the standard R `$` notation and the name we assigned the inputs in the `ui`. Hence, we can replace `"Primary Education"` with `input$selected_type`

#### Creating the plot
To ensure that the whole analysis is reactive to the user's input we need to put it in curly brackets itself and output the whole operation to an object named `biggestPlot` which we called in the `ui` section. We do this using the `renderPlot()` function and send it to another temporary object called `output`.

```{r eval=FALSE}
server <- function(input, output) {

    output$biggestPlot <- renderPlot({

      X8_sa_enrolment_by_school %>%
        filter(School_Type == input$selected_type) %>%
        slice_max(order_by = Year_2021, n = 12) %>%
        ggplot(aes(x = Year_2021, 
                   y = reorder(School_Name, Year_2021), 
                   fill = School_Type)) + 
        geom_col()
        
    })
}
```

### Wrapping it all up

There are actually three other things to remember to put in an `app.R` file.

1) The final line which is `shinyApp(ui = ui, server = server)`.
2) All the `library()` calls needed to create all the functions that are used in the app. These should be above the `ui` section so that they are available to all the code below.
3) The data that is used. This should also be placed above the `ui` section because it is used in both the `ui` and `server` sections.

Putting it all together we get:

```{r eval=FALSE}
library(shiny)
library(readr)
library(tidyverse)

X8_sa_enrolment_by_school <- read_csv("https://data.sa.gov.au/data/dataset/d385385a-6d79-4ef7-ab0f-0cfc583469f6/resource/7223f681-d340-4fc9-b92b-403d158f521f/download/8-sa-enrolment-by-school-.csv")

ui <- fluidPage(

    titlePanel("Biggest schools by enrollment"),

    # Sidebar with a slider input for number of bins 
    sidebarLayout(
        sidebarPanel(
          radioButtons("selected_type",
                        "Type of school",
                       unique(X8_sa_enrolment_by_school$School_Type)
          )
        ),

        mainPanel(
           plotOutput("biggestPlot")
        )
    )
)

server <- function(input, output) {

    output$biggestPlot <- renderPlot({

      X8_sa_enrolment_by_school %>%
        filter(School_Type == input$selected_type) %>%
        slice_max(order_by = Year_2021, n = 12) %>%
        ggplot(aes(x = Year_2021, 
                   y = reorder(School_Name, Year_2021), 
                   fill = School_Type)) + 
        geom_col(show.legend = FALSE) + 
        labs(title = "Biggest schools 2021", x = "Enrollments 2021", y = NULL)
        
    })
}

# Run the application 
shinyApp(ui = ui, server = server)

```

## Check it out
You can run the app by clicking on the "Run App" button in the ribbon at the top of the top left panel in RStudio. This will launch a browser that runs the app localy on your computer. It should look like the app here: https://ofchurches.shinyapps.io/biggest_schools/.

